---
title: "05 — Network Analysis: Disease Intersection, Networks & Functional Enrichment"
author: "Edgard Uriel López Chablé"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Overview

1. Load disease targets from **GeneCards**
2. Venn diagram of compound targets ∩ disease targets
3. Compound–target network **(Network B)**
4. STRING protein–protein interaction network **(Network C)**
5. Hub gene subgraph **(Network D)**
6. Functional enrichment: GO, KEGG, Reactome + dotplots + pathway maps + tripartite cnetplot **(Network E)**

```{r example-flag}
USE_EXAMPLE_DATA <- TRUE
EXAMPLE_DIR      <- "example_data"
```

```{r libraries}
library(tidyverse)
library(httr)
library(jsonlite)
library(igraph)
library(ggraph)
library(ggVennDiagram)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ReactomePA)
library(pathview)
library(enrichplot)
library(magick)
```

---

# Step 1 — Load target predictions from Step 04

```{r load-targets}
df_targets <- read_csv("04_target_prediction/df_targets.csv",
                        show_col_types = FALSE)

cat("Compound–target interactions:", nrow(df_targets), "\n")
cat("Unique compounds:", n_distinct(df_targets$Name), "\n")
cat("Unique targets:", n_distinct(df_targets$UniProtID), "\n")
```

---

# Step 2 — Disease Targets from GeneCards

## Manual step — Download GeneCards lists

> **Action required (skip when using example data):**
>
> 1. Go to https://www.genecards.org/
> 2. Search for each disease/phenotype term (e.g., "Hypertension", "Inflammation")
> 3. Download the results table as CSV
> 4. Save as `GeneCards<Term>.csv` (e.g., `GeneCardsHypertension.csv`) in `05_network_analysis/genecards/`

## ⚠️ Header check

```{r genecards-headers}
if (USE_EXAMPLE_DATA) {
  gc_dir <- file.path(EXAMPLE_DIR, "05_genecards")
} else {
  gc_dir <- "05_network_analysis/genecards"
}

gc_files <- list.files(gc_dir, pattern = "\\.csv$", full.names = TRUE)
cat("GeneCards files found:", length(gc_files), "\n")

if (length(gc_files) > 0) {
  sample_gc <- read_csv(gc_files[1], show_col_types = FALSE, n_max = 2)
  cat("\nHeaders in", basename(gc_files[1]), ":\n")
  print(colnames(sample_gc))
}
```

> Expected GeneCards headers:
>
> | Column | Notes |
> |--------|-------|
> | `Gene Symbol` | HGNC symbol |
> | `Uniprot ID` | UniProt accession (note lowercase 'n') |
> | `Relevance score` | GeneCards score — filter > 5 to remove low-confidence entries |

```{r genecards-load}
df_disease <- gc_files %>%
  map_dfr(~read_csv(.x, show_col_types = FALSE)) %>%
  filter(`Relevance score` > 5) %>%
  distinct(`Gene Symbol`, .keep_all = TRUE)

cat("Unique disease targets after filter:", nrow(df_disease), "\n")
dir.create("05_network_analysis/output", recursive = TRUE, showWarnings = FALSE)
write_csv(df_disease, "05_network_analysis/output/df_disease_targets.csv")
```

---

# Step 3 — Venn Diagram

```{r venn}
ids_compounds <- df_targets %>% pull(UniProtID) %>% unique() %>% na.omit()
ids_disease   <- df_disease %>% pull(`Uniprot ID`) %>% unique() %>% na.omit()

venn_list <- list(
  "Compound Targets" = ids_compounds,
  "Disease Targets"  = ids_disease
)

plot_venn <- ggVennDiagram(
  venn_list,
  category.names  = c("Compound\nTargets", "Disease\nTargets"),
  set_color       = "black",
  set_size        = 4,
  label           = "both",
  label_size      = 4,
  label_alpha     = 0
) +
  scale_fill_gradient(low = "#F4FAFE", high = "#4981BF") +
  scale_x_continuous(expand = expansion(mult = c(0.4, 0.4))) +
  theme_void() +
  labs(title    = "Compound–Disease Target Intersection",
       subtitle = "Compound targets vs. GeneCards disease targets") +
  theme(plot.title    = element_text(size = 18, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 13, hjust = 0.5),
        legend.position = "none")

print(plot_venn)

ggsave("05_network_analysis/output/Venn_Compounds_vs_Disease.tiff",
       plot = plot_venn, width = 10, height = 8, units = "in",
       dpi = 600, compression = "lzw", bg = "white")
ggsave("05_network_analysis/output/Venn_Compounds_vs_Disease.pdf",
       plot = plot_venn, width = 10, height = 8, units = "in",
       bg = "white", device = cairo_pdf)
```

---

# Step 4 — Build the Final Network Table

```{r network-table}
# Drop Gene Symbol from disease table to avoid column conflict after join
# (we use the Gene Symbol already present in df_targets from the prediction server)
network_final <- df_targets %>%
  inner_join(
    df_disease %>% dplyr::select(-`Gene Symbol`),
    by = c("UniProtID" = "Uniprot ID")
  )

cat("Interactions in intersection:", nrow(network_final), "\n")
cat("Unique targets:", n_distinct(network_final$UniProtID), "\n")

write_csv(network_final, "05_network_analysis/output/network_final.csv")
```

---

# Step 5 — Network B: Compound–Target Network

```{r network-b}
# ── Aesthetic parameters — adjust as needed ───────────────────────────────────
color_compounds <- "#5c5cff"   # blue
color_genes     <- "#ffcc00"   # yellow/gold

# ── Build graph ───────────────────────────────────────────────────────────────
edges_b <- network_final %>%
  dplyr::select(Source = Name, Target = `Gene Symbol`) %>%
  distinct()

graph_b <- graph_from_data_frame(d = edges_b, directed = FALSE)
V(graph_b)$degree <- degree(graph_b)
compound_names    <- unique(network_final$Name)
V(graph_b)$type   <- if_else(V(graph_b)$name %in% compound_names, "Compound", "Gene")

# ── Plot ──────────────────────────────────────────────────────────────────────
set.seed(42)

plot_b <- ggraph(graph_b, layout = "kk") +
  geom_edge_link(color = "grey85", alpha = 0.4, width = 0.3) +
  geom_node_point(aes(shape = type, color = type, size = degree)) +
  scale_shape_manual(values = c("Compound" = 18, "Gene" = 19)) +
  scale_color_manual(values = c("Compound" = color_compounds, "Gene" = color_genes)) +
  scale_size_continuous(range = c(2, 9)) +
  geom_node_text(
    aes(filter = (type == "Gene" & degree > 2), label = name),
    repel = TRUE, size = 3, color = "black", fontface = "bold",
    max.overlaps = 50, force = 3, box.padding = 0.8, point.padding = 0.4,
    bg.colour = "white", bg.r = 0.15
  ) +
  geom_node_text(
    aes(filter = (type == "Compound"), label = name),
    repel = TRUE, size = 3.5, color = "darkblue", fontface = "bold",
    max.overlaps = 50, force = 5, box.padding = 1.2, point.padding = 0.5,
    bg.colour = "white", bg.r = 0.15
  ) +
  theme_graph() +
  labs(title = "Compound–Target Network",
       shape = "Node Type", color = "Node Type", size = "Degree") +
  theme(legend.position = "right")

print(plot_b)

ggsave("05_network_analysis/output/Network_B_Compound_Target.tiff",
       plot = plot_b, width = 12, height = 10, units = "in",
       dpi = 600, compression = "lzw", bg = "white")
ggsave("05_network_analysis/output/Network_B_Compound_Target.pdf",
       plot = plot_b, width = 12, height = 10, units = "in",
       bg = "white", device = cairo_pdf)
```

---

# Step 6 — Network C: PPI from STRING

```{r network-c}
# ── STRING REST API (avoids downloading the full ~80 MB proteome links file) ──
# Only fetches interactions for your specific gene list — much faster and reliable.
library(httr)
library(jsonlite)

gene_symbols <- network_final %>%
  pull(`Gene Symbol`) %>%
  unique() %>%
  na.omit()

# required_score: 400 = medium, 700 = high, 900 = very high
# POST is required for multiple identifiers — GET double-encodes the \r separator
string_response <- POST(
  "https://string-db.org/api/json/network",
  body = list(
    identifiers     = paste(gene_symbols, collapse = "\r"),
    species         = "9606",
    required_score  = "700",
    caller_identity = "network_pharmacology_cookbook"
  ),
  encode = "form"
)

if (http_error(string_response)) {
  stop("STRING API request failed: ", status_code(string_response))
}

raw_text <- content(string_response, "text", encoding = "UTF-8")

# Empty response = no interactions found above the score threshold
if (nchar(trimws(raw_text)) == 0 || raw_text == "[]") {
  warning("STRING returned no interactions. Try lowering required_score to 400.")
  ppi_network <- tibble(Source = character(), Target = character(),
                        combined_score = numeric())
} else {
  # STRING REST API returns score in 0–1 scale; required_score parameter
  # already pre-filters at the requested threshold (700/1000 = 0.7).
  # No additional filter needed here.
  ppi_network <- fromJSON(raw_text) %>%
    as_tibble() %>%
    dplyr::select(Source = preferredName_A, Target = preferredName_B,
                  combined_score = score) %>%
    distinct()
}

cat("PPI interactions extracted:", nrow(ppi_network), "\n")
write_csv(ppi_network, "05_network_analysis/output/PPI_EdgeList.csv")

# ── Plot ──────────────────────────────────────────────────────────────────────
set.seed(123)
graph_c           <- graph_from_data_frame(d = ppi_network, directed = FALSE)
V(graph_c)$degree <- degree(graph_c)

plot_c <- ggraph(graph_c, layout = "kk") +
  geom_edge_link(color = "grey88", alpha = 0.5, width = 0.3) +
  geom_node_point(aes(size = degree, color = degree)) +
  scale_color_gradient(low = "#ff9933", high = "#990000") +
  scale_size_continuous(range = c(3, 9)) +
  geom_node_text(
    aes(label = if_else(degree > 1, name, "")),
    repel = TRUE, size = 3, color = "black", fontface = "bold",
    max.overlaps = 30, bg.colour = "white", bg.r = 0.15
  ) +
  theme_graph() +
  labs(title = "Protein–Protein Interaction Network (PPI)",
       color = "Degree", size = "Degree") +
  theme(legend.position = "right")

print(plot_c)

ggsave("05_network_analysis/output/Network_C_PPI.tiff",
       plot = plot_c, width = 11, height = 9, units = "in",
       dpi = 600, compression = "lzw", bg = "white")
ggsave("05_network_analysis/output/Network_C_PPI.pdf",
       plot = plot_c, width = 11, height = 9, units = "in",
       bg = "white", device = cairo_pdf)
```

---

# Step 7 — Network Topology Metrics

> **Why report topology?** Reviewers of *Frontiers in Pharmacology*, *Phytomedicine*,
> and similar journals routinely ask for a network topology table. It confirms that
> your PPI network has "small-world" properties (high clustering, short paths) typical
> of biological networks, and distinguishes it from a random graph.

```{r topology}
topology_metrics <- tibble(
  Metric  = c("Nodes", "Edges", "Average degree",
               "Network density", "Clustering coefficient",
               "Average path length", "Diameter"),
  Value = c(
    vcount(graph_c),
    ecount(graph_c),
    round(mean(degree(graph_c)), 3),
    round(edge_density(graph_c), 4),
    round(transitivity(graph_c, type = "average"), 3),
    round(mean_distance(graph_c, directed = FALSE), 3),
    diameter(graph_c, directed = FALSE)
  )
)

write_csv(topology_metrics, "05_network_analysis/output/Network_Topology.csv")
print(topology_metrics)
```

---

# Step 8 — Network D: Hub Gene Subgraph (Multi-Centrality)

> **Why multiple centrality measures?**
> Degree centrality counts how many proteins a node is connected to — a high-degree
> protein interacts with many others. But **betweenness centrality** identifies proteins
> that act as *bridges* between clusters (knock them out and the network falls apart).
> **Closeness centrality** measures how quickly a signal can propagate from that protein
> to the rest of the network. Using all three gives a more robust hub definition than
> degree alone, which is what CytoHubba (the Cytoscape plugin) does internally.
>
> A gene that ranks high on all three measures is the strongest candidate for docking:
> it is both highly connected AND structurally critical to the network.

```{r network-d}
n_hubs <- min(10, vcount(graph_c))

# ── Compute three centrality measures ────────────────────────────────────────
centrality_table <- tibble(
  Gene        = V(graph_c)$name,
  Degree      = degree(graph_c),
  Betweenness = round(betweenness(graph_c, normalized = TRUE), 4),
  Closeness   = round(closeness(graph_c, normalized = TRUE), 4)
) %>%
  # Composite score: average rank across the three measures
  mutate(
    Rank_Degree      = rank(-Degree),
    Rank_Betweenness = rank(-Betweenness),
    Rank_Closeness   = rank(-Closeness),
    Composite_Score  = round((Rank_Degree + Rank_Betweenness + Rank_Closeness) / 3, 2)
  ) %>%
  arrange(Composite_Score) %>%
  dplyr::select(Gene, Degree, Betweenness, Closeness, Composite_Score)

write_csv(centrality_table, "05_network_analysis/output/HubGene_Table.csv")
cat("=== Hub Gene Centrality Table ===\n")
print(centrality_table)

# Top hubs by composite score for the subgraph
top_hub_names <- centrality_table %>% slice_head(n = n_hubs) %>% pull(Gene)
hub_vids      <- which(V(graph_c)$name %in% top_hub_names)
graph_hubs    <- induced_subgraph(graph_c, vids = hub_vids)

V(graph_hubs)$degree      <- degree(graph_hubs)
V(graph_hubs)$betweenness <- betweenness(graph_hubs, normalized = TRUE)

# ── Plot: node size = degree, color = betweenness ────────────────────────────
plot_d <- ggraph(graph_hubs, layout = "circle") +
  geom_edge_link(color = "darkslategray4", alpha = 0.6, width = 0.8) +
  geom_node_point(aes(size = degree, color = betweenness)) +
  scale_color_gradient(low = "#e63946", high = "#7a0010",
                       name = "Betweenness\n(normalized)") +
  scale_size_continuous(range = c(8, 16), name = "Degree") +
  geom_node_text(aes(label = name), repel = TRUE, size = 4.5,
                 color = "black", fontface = "bold",
                 bg.colour = "white", bg.r = 0.2) +
  theme_graph() +
  labs(title    = paste("Top", n_hubs, "Hub Genes — PPI Network"),
       subtitle = "Node size = Degree  |  Color = Betweenness centrality") +
  theme(legend.position = "right")

print(plot_d)

ggsave("05_network_analysis/output/Network_D_HubGenes.tiff",
       plot = plot_d, width = 9, height = 8, units = "in",
       dpi = 600, compression = "lzw", bg = "white")
ggsave("05_network_analysis/output/Network_D_HubGenes.pdf",
       plot = plot_d, width = 9, height = 8, units = "in",
       bg = "white", device = cairo_pdf)
```

---

# Step 8 — Functional Enrichment

## 8a. Prepare data and translate IDs

```{r enrich-prep}
# Aggregate by target
target_summary <- network_final %>%
  filter(Probability > 0) %>%
  group_by(UniProtID) %>%
  summarise(
    n_compounds = n(),
    max_prob    = max(Probability),
    compounds   = paste(unique(Name), collapse = ", "),
    gene_symbol = dplyr::first(`Gene Symbol`)
  ) %>%
  ungroup()

# UniProt → ENTREZID (required by KEGG and GO)
id_mapping <- bitr(target_summary$UniProtID,
                   fromType = "UNIPROT",
                   toType   = c("ENTREZID", "SYMBOL"),
                   OrgDb    = org.Hs.eg.db)

enrichment_data <- target_summary %>%
  inner_join(id_mapping, by = c("UniProtID" = "UNIPROT")) %>%
  mutate(label_tripartite = paste0(SYMBOL, " (", compounds, ")"))

cat("Genes ready for enrichment:", nrow(enrichment_data), "\n")
```

## 8b. Run GO, KEGG, and Reactome

```{r enrichment-run}
ego    <- enrichGO(gene = enrichment_data$ENTREZID, OrgDb = org.Hs.eg.db,
                   ont = "BP", readable = TRUE)
ekegg  <- enrichKEGG(gene = enrichment_data$ENTREZID, organism = "hsa")
ereact <- enrichPathway(gene = enrichment_data$ENTREZID, organism = "human",
                        readable = TRUE)

write_csv(as.data.frame(ego),    "05_network_analysis/output/Results_GO.csv")
write_csv(as.data.frame(ekegg),  "05_network_analysis/output/Results_KEGG.csv")
write_csv(as.data.frame(ereact), "05_network_analysis/output/Results_Reactome.csv")
cat("Enrichment tables saved.\n")
```

## 8c. Dotplots

```{r dotplots}
save_plot <- function(p, filename) {
  ggsave(paste0("05_network_analysis/output/", filename, ".tiff"),
         plot = p, width = 10, height = 8, units = "in",
         dpi = 600, compression = "lzw", bg = "white")
  ggsave(paste0("05_network_analysis/output/", filename, ".pdf"),
         plot = p, width = 10, height = 8, units = "in",
         bg = "white", device = cairo_pdf)
}

plot_go       <- dotplot(ego,    showCategory = 15, title = "GO Biological Process")
plot_kegg     <- dotplot(ekegg,  showCategory = 15, title = "KEGG Pathways")
plot_reactome <- dotplot(ereact, showCategory = 15, title = "Reactome Pathways")

save_plot(plot_go,       "Dotplot_GO")
save_plot(plot_kegg,     "Dotplot_KEGG")
save_plot(plot_reactome, "Dotplot_Reactome")

print(plot_go)
print(plot_kegg)
```

## 8d. KEGG Pathway Maps (pathview)

```{r pathview}
gene_vector_kegg        <- enrichment_data$max_prob
names(gene_vector_kegg) <- enrichment_data$ENTREZID

top_kegg_ids   <- as.data.frame(ekegg) %>% head(10) %>% pull(ID)
pathview_dir   <- "05_network_analysis/output/pathview"
dir.create(pathview_dir, recursive = TRUE, showWarnings = FALSE)

# pathview() has no output directory parameter — it always writes to the
# current working directory. We run it, then move every generated file to
# the output folder so the cookbook root stays clean.
for (kegg_id in top_kegg_ids) {
  tryCatch(
    pathview(gene.data  = gene_vector_kegg,
             pathway.id = kegg_id,
             species    = "hsa",
             limit      = list(gene = c(0, 1)),
             low = "white", mid = "yellow", high = "red"),
    error = function(e) message("Error in pathway: ", kegg_id)
  )
}

# Move all pathview-generated files (PNG, XML, PDF) to output/pathview/
generated <- list.files(
  pattern = paste0("(", paste(top_kegg_ids, collapse = "|"), ").*\\.(png|xml|pdf)$")
)
if (length(generated) > 0) {
  file.rename(generated, file.path(pathview_dir, generated))
  cat("Moved", length(generated), "pathview files →", pathview_dir, "\n")
}

# Compile all .pathview.png files into a single PDF
pathway_pngs <- list.files(pathview_dir, pattern = "\\.pathview\\.png$",
                            full.names = TRUE)
if (length(pathway_pngs) > 0) {
  pdf("05_network_analysis/output/KEGG_Pathway_Maps.pdf", width = 14, height = 10)
  for (img_path in pathway_pngs) {
    plot(as.raster(image_read(img_path)))
    title(main = paste("Pathway:", str_extract(basename(img_path), "hsa\\d+")),
          cex.main = 2, col.main = "darkblue")
  }
  dev.off()
  cat("KEGG pathway maps compiled →",
      "05_network_analysis/output/KEGG_Pathway_Maps.pdf\n")
}
```

## 8e. Network E: Tripartite Reactome cnetplot

```{r cnetplot}
gene_vector_cnet        <- enrichment_data$max_prob
names(gene_vector_cnet) <- enrichment_data$SYMBOL

plot_e <- cnetplot(ereact,
                   showCategory   = 5,
                   foldChange     = gene_vector_cnet,
                   node_label     = "all",
                   color_category = "black")

# Embed compound names in gene labels: "SYMBOL (compound1, compound2)"
plot_e$data$label[plot_e$data$label %in% enrichment_data$SYMBOL] <-
  enrichment_data$label_tripartite[
    match(plot_e$data$label[plot_e$data$label %in% enrichment_data$SYMBOL],
          enrichment_data$SYMBOL)
  ]

plot_e_final <- plot_e +
  scale_color_gradient(name = "Max Prediction\nProbability",
                       low = "#ffcc00", high = "#cc0000") +
  theme_void() +
  labs(title    = "Tripartite Network — Reactome",
       subtitle = "Pathway → Target (Compounds predicted to interact)") +
  theme(legend.position = "right")

print(plot_e_final)
save_plot(plot_e_final, "Network_E_Tripartite_Reactome")
cat("All network analysis steps complete.\n")
```

---

# Output

| File | Description |
|------|-------------|
| `df_disease_targets.csv` | GeneCards disease targets (merged) |
| `Venn_Compounds_vs_Disease.*` | Venn diagram (TIFF + PDF) |
| `network_final.csv` | Compound–target intersection table |
| `Network_B_Compound_Target.*` | Bipartite compound–target network |
| `PPI_EdgeList.csv` | STRING PPI edge list |
| `Network_Topology.csv` | Nodes, edges, density, clustering coefficient, path length |
| `Network_C_PPI.*` | Full PPI network |
| `Network_D_HubGenes.*` | Hub gene subgraph (size=degree, color=betweenness) |
| `HubGene_Table.csv` | Degree + betweenness + closeness + composite score |
| `Results_GO/KEGG/Reactome.csv` | Enrichment tables |
| `Dotplot_*.tiff/.pdf` | Enrichment dotplots |
| `KEGG_Pathway_Maps.pdf` | Pathview maps compiled |
| `Network_E_Tripartite_Reactome.*` | Reactome cnetplot |

→ Feeds into **Step 06 (Pre-Docking Analysis)**
