---
title: "06 — Pre-Docking Analysis: Heatmaps, UpSet Plots & Candidate Prioritization"
author: "Edgard Uriel López Chablé"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Overview

Molecular docking is computationally expensive. This step identifies the highest-priority compound–target pairs **before** spending GPU time.

Three questions answered:

1. **Which targets are hit by the most compounds?** → Hub target bar chart
2. **Which compounds share the most targets?** → UpSet plot (reveals synergistic clusters)
3. **What is the predicted binding probability per compound–target pair?** → Heatmap (core prioritization tool)

Compounds with high binding probability AND targeting multiple disease-relevant proteins (high degree) are the strongest docking candidates.

```{r example-flag}
USE_EXAMPLE_DATA <- TRUE
```

```{r libraries}
library(tidyverse)
library(pheatmap)
library(UpSetR)
library(RColorBrewer)
library(igraph)
library(grid)
```

---

# Load data

```{r load}
network_final <- read_csv("05_network_analysis/output/network_final.csv",
                           show_col_types = FALSE)

cat("Interactions:", nrow(network_final), "\n")
cat("Compounds:   ", n_distinct(network_final$Name), "\n")
cat("Targets:     ", n_distinct(network_final$`Gene Symbol`), "\n")
```

---

# Step 1 — Hub Target Bar Chart

Ranks targets by number of compound connections. High-degree targets are the primary docking candidates: multiple compounds from your extract are predicted to engage them, suggesting a convergent pharmacological mechanism.

```{r hub-bar}
dir.create("06_predocking_analysis/output", recursive = TRUE, showWarnings = FALSE)

# Build bipartite graph to compute degree per node
edges_bipartite <- network_final %>%
  dplyr::select(Name, `Gene Symbol`) %>%
  distinct()

graph_bipartite  <- graph_from_data_frame(edges_bipartite, directed = FALSE)
all_degrees      <- degree(graph_bipartite)
compound_names   <- unique(network_final$Name)
target_degrees   <- all_degrees[!names(all_degrees) %in% compound_names]

n_top <- 10  # Number of hub targets to display

hub_targets <- tibble(
  Target = names(target_degrees),
  Degree = as.integer(target_degrees)
) %>%
  arrange(desc(Degree)) %>%
  slice_head(n = n_top)

# Plot
plot_hub_bar <- ggplot(hub_targets, aes(x = reorder(Target, Degree), y = Degree)) +
  geom_bar(stat = "identity", fill = "#3498db", width = 0.7) +
  coord_flip() +
  labs(
    title    = paste("Top", n_top, "Hub Targets"),
    subtitle = "Number of compounds predicted to interact with each target",
    x        = "Target (Gene Symbol)",
    y        = "Compound Interactions (Degree)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

print(plot_hub_bar)

ggsave("06_predocking_analysis/output/Hub_Targets_BarChart.tiff",
       plot = plot_hub_bar, width = 10, height = 7, units = "in",
       dpi = 600, compression = "lzw", bg = "white")
ggsave("06_predocking_analysis/output/Hub_Targets_BarChart.pdf",
       plot = plot_hub_bar, width = 10, height = 7, units = "in",
       bg = "white", device = cairo_pdf)

write_csv(hub_targets, "06_predocking_analysis/output/Hub_Targets_Ranked.csv")
cat("Hub target bar chart saved.\n")
```

---

# Step 2 — UpSet Plot: Shared Targets Between Compounds

Shows target overlaps across compounds. Large intersections indicate cooperative action — these shared targets are high-value docking candidates.

```{r upset}
upset_list <- network_final %>%
  filter(Probability > 0) %>%
  group_by(Name) %>%
  summarise(Targets = list(`Gene Symbol`)) %>%
  deframe()

pdf("06_predocking_analysis/output/UpSet_Compound_Target_Overlap.pdf",
    width = 16, height = 10)

upset(
  fromList(upset_list),
  nsets       = min(10, length(upset_list)),
  nintersects = 20,
  order.by    = "freq",
  main.bar.color = "#2c3e50",
  sets.bar.color = "#e74c3c",
  text.scale     = c(1.3, 1.3, 1, 1, 1.2, 1.2)
)

grid.text("Target Intersection by Compound",
          x = 0.65, y = 0.95,
          gp = gpar(fontsize = 20, fontface = "bold"))

dev.off()
cat("UpSet plot saved.\n")
```

---

# Step 3 — Heatmap: Compound × Target Binding Probability

Rows = compounds, columns = hub targets, color = predicted binding probability.

**How to read it:**
- Dark red = high predicted affinity → prioritize for docking
- White = no predicted interaction
- Hierarchical clustering groups compounds with similar target profiles

```{r heatmap}
# Use hub targets as columns
top_target_names <- hub_targets$Target

# Build probability matrix
heatmap_long <- network_final %>%
  filter(`Gene Symbol` %in% top_target_names) %>%
  dplyr::select(Name, `Gene Symbol`, Probability) %>%
  distinct(Name, `Gene Symbol`, .keep_all = TRUE)

heatmap_wide <- heatmap_long %>%
  pivot_wider(names_from = `Gene Symbol`, values_from = Probability,
              values_fill = 0)

# Convert to named matrix
heatmap_matrix           <- as.data.frame(heatmap_wide)
rownames(heatmap_matrix) <- heatmap_matrix$Name
heatmap_matrix$Name      <- NULL

# Reorder columns to match hub ranking
col_order       <- intersect(top_target_names, colnames(heatmap_matrix))
heatmap_matrix  <- heatmap_matrix[, col_order, drop = FALSE]

pdf("06_predocking_analysis/output/Heatmap_Compound_Target_Probability.pdf",
    width  = max(8, ncol(heatmap_matrix) * 0.9 + 2),
    height = max(6, nrow(heatmap_matrix) * 0.5 + 2))

pheatmap(
  heatmap_matrix,
  color           = colorRampPalette(c("white", "#f39c12", "#c0392b"))(50),
  display_numbers = TRUE,
  number_format   = "%.2f",
  fontsize_number = 9,
  cluster_rows    = TRUE,
  cluster_cols    = TRUE,
  main            = "Compound–Target Binding Probability",
  angle_col       = 45,
  fontsize_row    = 10,
  fontsize_col    = 10
)

dev.off()
cat("Heatmap saved.\n")
```

---

# Step 4 — Candidate Prioritization Table

Combines hub degree, binding probability, and compound hit count into a ranked table. This is the direct input for selecting compound–target pairs to submit to molecular docking.

```{r prioritize}
docking_candidates <- network_final %>%
  filter(`Gene Symbol` %in% top_target_names, Probability > 0) %>%
  group_by(`Gene Symbol`, UniProtID) %>%
  summarise(
    n_compounds     = n(),
    max_probability = max(Probability),
    mean_probability = round(mean(Probability), 3),
    top_compounds   = paste(Name[order(-Probability)][1:min(3, n())], collapse = ", "),
    .groups = "drop"
  ) %>%
  left_join(
    hub_targets %>% dplyr::select(`Gene Symbol` = Target, Degree),
    by = "Gene Symbol"
  ) %>%
  arrange(desc(Degree), desc(max_probability)) %>%
  dplyr::rename(
    Gene       = `Gene Symbol`,
    Hub_Degree = Degree,
    UniProt_ID = UniProtID
  )

write_csv(docking_candidates, "06_predocking_analysis/output/Docking_Candidates_Ranked.csv")

cat("=== Top Docking Candidates ===\n")
print(docking_candidates %>%
        dplyr::select(Gene, Hub_Degree, n_compounds, max_probability, top_compounds))
```

---

# Step 5 — Drug-Gene Interaction Cross-Validation (DGIdb)

> **Why cross-validate with known drugs?**
> If an approved drug already targets one of your hub genes for the same disease you
> are studying, that is strong pharmacological evidence that the target is relevant.
> It does *not* mean your compounds are redundant — a plant phenolic acting on the
> same target may have a better safety profile, fewer off-target effects, or a
> complementary binding pose. Reviewers see this as evidence that your target
> selection is biologically justified, not arbitrary.
>
> We use **DGIdb** (Drug Gene Interaction Database, https://dgidb.org) because it
> aggregates DrugBank, ChEMBL, TTD, and other sources and has a free REST API that
> does not require registration. This step runs automatically.

```{r dgidb}
# ── DGIdb v4+ uses GraphQL — the old REST v2 endpoint is deprecated ───────────
# GraphQL lets us request exactly the fields we need in a single query.

gql_query <- sprintf('
{
  genes(names: [%s]) {
    nodes {
      name
      interactions {
        drug { name approved }
        interactionScore
        interactionTypes { type directionality }
        sources { sourceDbName }
      }
    }
  }
}',
paste0('"', top_target_names, '"', collapse = ", ")
)

dgidb_response <- POST(
  "https://dgidb.org/api/graphql",
  body    = toJSON(list(query = gql_query), auto_unbox = TRUE),
  add_headers(`Content-Type` = "application/json"),
  encode  = "raw"
)

if (http_error(dgidb_response)) {
  warning("DGIdb GraphQL unavailable. Check https://dgidb.org manually.")
  drug_validation <- tibble(Gene = character(), Drug = character(),
                             Approved = logical(), Interaction_Type = character(),
                             Source = character())
} else {
  parsed <- fromJSON(content(dgidb_response, "text", encoding = "UTF-8"),
                     simplifyVector = FALSE)

  nodes <- parsed$data$genes$nodes

  drug_validation <- map_dfr(nodes, function(gene_node) {
    ixns <- gene_node$interactions
    if (length(ixns) == 0) return(NULL)
    map_dfr(ixns, function(ix) {
      tibble(
        Gene             = gene_node$name,
        Drug             = ix$drug$name %||% NA_character_,
        Approved         = ix$drug$approved %||% NA,
        Interaction_Type = if (length(ix$interactionTypes) > 0)
                             ix$interactionTypes[[1]]$type else NA_character_,
        Source           = if (length(ix$sources) > 0)
                             paste(map_chr(ix$sources, "sourceDbName"),
                                   collapse = "; ") else NA_character_
      )
    })
  }) %>%
    filter(!is.na(Drug)) %>%
    distinct()

  cat("Drug–gene interactions found:", nrow(drug_validation), "\n")
  cat("Hub genes with known drugs:", n_distinct(drug_validation$Gene),
      "/", length(top_target_names), "\n")

  novel_targets <- setdiff(top_target_names, unique(drug_validation$Gene))
  if (length(novel_targets) > 0) {
    cat("\nHub genes with no known drug interactions (potentially novel):\n")
    cat(paste(novel_targets, collapse = ", "), "\n")
  }

  write_csv(drug_validation,
            "06_predocking_analysis/output/DrugGene_Validation.csv")

  drug_summary <- drug_validation %>%
    group_by(Gene) %>%
    summarise(
      n_drugs  = n_distinct(Drug),
      approved = sum(Approved == TRUE, na.rm = TRUE),
      drugs    = paste(unique(Drug)[seq_len(min(3, n_distinct(Drug)))],
                       collapse = "; "),
      .groups  = "drop"
    ) %>%
    arrange(desc(n_drugs))

  print(drug_summary)
}
```

---

# Step 6 — GOChord Diagram (Optional)

Chord diagram connecting KEGG pathways to genes. Useful for visualizing which hub genes participate in multiple pathways simultaneously.

```{r gochord, eval=FALSE}
library(GOplot)
library(clusterProfiler)

# Load enrichment object from Step 05
# ekegg <- readRDS("05_network_analysis/output/ekegg.rds")  # if you saved it

ekegg_readable <- setReadable(ekegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
df_kegg        <- as.data.frame(ekegg_readable)
top5           <- df_kegg[1:5, ]
all_genes      <- unique(unlist(strsplit(top5$geneID, "/")))

chord_matrix           <- matrix(0, nrow = length(all_genes), ncol = nrow(top5))
rownames(chord_matrix) <- all_genes
colnames(chord_matrix) <- top5$Description

for (i in seq_len(nrow(top5))) {
  genes_in_path <- unlist(strsplit(top5$geneID[i], "/"))
  chord_matrix[genes_in_path, i] <- 1
}

chord_matrix <- cbind(chord_matrix,
                      logFC = seq(-1, 1, length.out = nrow(chord_matrix)))

pdf("06_predocking_analysis/output/GOChord_KEGG_Top5.pdf", width = 14, height = 14)
GOChord(chord_matrix,
        space         = 0.03,
        gene.order    = "logFC",
        gene.space    = 0.28,
        gene.size     = 4,
        process.label = 10,
        ribbon.col    = brewer.pal(5, "Set2"),
        lfc.col       = c("#E74C3C", "white", "#3498DB"),
        border.size   = 0.1)
dev.off()
cat("GOChord saved.\n")
```

---

# Output Summary

| File | Description |
|------|-------------|
| `Hub_Targets_BarChart.*` | Bar chart of top hub targets by compound degree |
| `Hub_Targets_Ranked.csv` | Ranked hub target table |
| `UpSet_Compound_Target_Overlap.pdf` | Shared target intersections between compounds |
| `Heatmap_Compound_Target_Probability.pdf` | Binding probability matrix (core prioritization) |
| `Docking_Candidates_Ranked.csv` | **Final ranked list for molecular docking** |
| `DrugGene_Validation.csv` | Known drugs per hub gene (DGIdb/DrugBank) |
| `GOChord_KEGG_Top5.pdf` | Gene–pathway chord diagram (optional) |

---

# Next Steps: Molecular Docking & MD

> **[ In development ]**
>
> Use `Docking_Candidates_Ranked.csv` as input for:
>
> 1. **Structure preparation** — Download protein structures from PDB (https://www.rcsb.org/), prepare with AutoDockTools or GNINA
> 2. **Ligand preparation** — Generate 3D conformers from SMILES (RDKit, OpenBabel)
> 3. **Molecular docking** — AutoDock Vina or GNINA
> 4. **Binding pose analysis** — PyMOL, PLIP
> 5. **Molecular dynamics (MD)** — GROMACS or AMBER for hit validation
