---
title: "01 — Data Cleaning: SMILES Retrieval & Compound Matrix Preparation"
author: "Edgard Uriel López Chablé"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))  # all paths relative to cookbook root
```

# Overview

This step prepares the compound matrix for ADME filtering. Two entry points:

- **Scenario A (GC-MS matrix):** Compound × sample abundance matrix with replicate columns per condition (format: `R<n>-<CONDITION>`, e.g. `R1-LEA-ET`, `R2-FLO-AQ`). This script retrieves SMILES, collapses replicates, and binarizes the matrix.
- **Scenario B (Curated list):** Table with compound names, PubChemCID, and SMILES already available. Skip directly to the export chunk.

Both scenarios produce: `compounds_ready.csv` with columns `Name`, `PubChemCID`, `SMILES`, and (Scenario A only) one binary column per condition.

```{r example-flag}
# ── Set to TRUE to run with the bundled example dataset ───────────────────────
# Set to FALSE and update the file paths in each load chunk for your own data.
USE_EXAMPLE_DATA <- TRUE
EXAMPLE_DIR      <- "example_data"
```

```{r libraries}
library(tidyverse)
library(webchem)
```

---

# Scenario A — GC-MS Abundance Matrix

## A1. Load data

```{r load-a}
if (USE_EXAMPLE_DATA) {
  # Example: 8 compounds × 5 conditions × 3 replicates
  # Columns: Name, R1-LEA-ET, R2-LEA-ET, ..., R3-STE-ET
  abundance_matrix    <- read_csv(file.path(EXAMPLE_DIR, "01_abundance_matrix.csv"),
                                   show_col_types = FALSE)
  compound_dictionary <- read_csv(file.path(EXAMPLE_DIR, "01_compound_dictionary.csv"),
                                   show_col_types = FALSE)
} else {
  # ── Replace with your actual file paths ───────────────────────────────────
  # abundance_matrix:    compound × sample matrix from GC-MS/metabolomics
  #   Required columns: Name (character) + replicate columns (R<n>-<CONDITION>)
  # compound_dictionary: master table with NAME and CAS columns
  abundance_matrix    <- read_csv("YOUR_ABUNDANCE_MATRIX.csv")
  compound_dictionary <- read_csv("YOUR_COMPOUND_DICTIONARY.csv")
}

cat("Compounds in matrix:", nrow(abundance_matrix), "\n")
cat("Sample columns:", ncol(abundance_matrix) - 1, "\n")
```

## A2. Retrieve SMILES from PubChem

We query PubChem **by compound name**, not by CAS, because a compound can have multiple registered CAS entries across databases, causing silent mismatches. PubChem name lookup resolves to a unique CID and canonical SMILES.

```{r smiles-pubchem}
# Helper: normalize compound names for reliable joining
normalize_name <- function(x) {
  x %>% tolower() %>%
    str_replace_all("[[:punct:]]", "") %>%
    str_replace_all("\\s+", "")
}

# Build a lookup dictionary: normalized name → CAS
dict_normalized <- compound_dictionary %>%
  select(NAME, CAS) %>%
  mutate(name_key = normalize_name(NAME)) %>%
  distinct(name_key, .keep_all = TRUE)

# Attach CAS to the abundance matrix via normalized name match
df_with_cas <- abundance_matrix %>%
  mutate(name_key = normalize_name(Name)) %>%
  left_join(dict_normalized, by = "name_key") %>%
  select(Name, CAS, everything(), -name_key, -NAME) %>%
  drop_na(CAS)

cat("Compounds with matched CAS:", nrow(df_with_cas), "\n")

# Query PubChem: CID by compound name
# NOTE: requires internet. Can take several minutes for large lists.
cids <- get_cid(df_with_cas$Name, from = "name", match = "first") %>%
  rename(Name = query) %>%
  drop_na(cid)

# Retrieve canonical SMILES for each CID
smiles_df <- pc_prop(cids$cid, properties = "SMILES") %>%
  rename(cid = CID, SMILES = SMILES)

# Integrate into matrix
df_with_smiles <- df_with_cas %>%
  left_join(cids,      by = "Name") %>%
  left_join(smiles_df, by = "cid") %>%
  drop_na(SMILES) %>%
  rename(PubChemCID = cid) %>%
  select(Name, CAS, PubChemCID, SMILES, everything()) %>%
  distinct()

cat("Compounds with SMILES:", nrow(df_with_smiles), "\n")
```

## A3. Collapse replicates and binarize by condition

Converts abundance-per-replicate → presence/absence-per-condition.

```{r binarize}
# NOTE: Adjust the regex if your replicate prefix differs from R<n>-
# Expected pattern: R1-CONDITION, R2-CONDITION, R3-CONDITION
# Examples: R1-LEA-ET, R2-FLO-AQ, R3-STE-ET

df_binary <- df_with_smiles %>%
  pivot_longer(
    cols      = matches("^R\\d+-"),    # replicate columns
    names_to  = "sample",
    values_to = "abundance"
  ) %>%
  mutate(condition = str_remove(sample, "^R\\d+-")) %>%  # e.g. R1-LEA-ET → LEA-ET
  group_by(Name, CAS, PubChemCID, SMILES, condition) %>%
  summarise(total = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  mutate(presence = if_else(total > 0, 1L, 0L)) %>%
  select(-total) %>%
  pivot_wider(names_from = condition, values_from = presence, values_fill = 0L)

cat("Binary matrix — compounds:", nrow(df_binary), "| conditions:", ncol(df_binary) - 4, "\n")
```

## A4. Add manually curated compounds (optional)

Some compounds of interest may be absent from GC-MS data (e.g., heat-labile compounds). Add them here with their known condition presence.

```{r manual-compounds}
# Adjust Name, CAS, PubChemCID, SMILES, and binary values per condition
manual_compounds <- tibble(
  Name       = c("Rutin"),
  CAS        = c("153-18-4"),
  PubChemCID = c(5280805L),
  SMILES     = c("OC[C@H]1O[C@@H](Oc2cc3c(=O)cc(-c4ccc(O)c(O)c4)oc3cc2O)[C@H](O)[C@@H](O)[C@@H]1O"),
  `LEA-ET`   = c(1L),
  `LEA-AQ`   = c(1L),
  `FLO-ET`   = c(0L),
  `FLO-AQ`   = c(0L),
  `STE-ET`   = c(0L)
)

df_binary_final <- bind_rows(
    df_binary        %>% mutate(PubChemCID = as.integer(PubChemCID)),
    manual_compounds %>% mutate(PubChemCID = as.integer(PubChemCID))
  ) %>%
  mutate(across(where(is.numeric), ~ replace_na(., 0L)))

cat("Final matrix with manual compounds:", nrow(df_binary_final), "compounds\n")
```

---

# Scenario B — Curated Compound List

Use this path when you already have compound names, PubChemCIDs, and SMILES (e.g., from a literature search or phytochemical database).

```{r scenario-b}
if (USE_EXAMPLE_DATA) {
  # Example: 8 phenolic compounds from Sphaeralcea angustifolia
  compounds_ready <- read_csv(file.path(EXAMPLE_DIR, "01_compounds_list.csv"),
                               show_col_types = FALSE)
} else {
  # ── Replace with your actual file path ────────────────────────────────────
  # Required columns: Name, CAS, PubChemCID, SMILES
  # Optional: condition binary columns (if coming from Scenario A)
  compounds_ready <- read_csv("YOUR_COMPOUNDS_LIST.csv")
}

cat("Compounds loaded (Scenario B):", nrow(compounds_ready), "\n")
print(compounds_ready)
```

---

# Export

```{r export}
# For Scenario A: export df_binary_final
# For Scenario B: the object is already called compounds_ready

# Uncomment the line matching your scenario:
# write_csv(df_binary_final, "01_data_cleaning/compounds_ready.csv")   # Scenario A
write_csv(compounds_ready, "01_data_cleaning/compounds_ready.csv")     # Scenario B

cat("Output: 01_data_cleaning/compounds_ready.csv\n")
```

---

# Output

| File | Description |
|------|-------------|
| `01_data_cleaning/compounds_ready.csv` | Compound table: `Name`, `PubChemCID`, `SMILES` ± binary condition columns |

→ Feeds into **Step 02 (ADME Filtering)**
