---
title: "04 — Target Prediction: File Convention & Consolidation"
author: "Edgard Uriel López Chablé"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Overview

Consolidates individual target prediction files (one per compound) into a single master data frame, then attaches compound metadata.

## Supported prediction servers

| Server | URL | Notes |
|--------|-----|-------|
| **SwissTargetPrediction** | https://www.swisstargetprediction.ch/ | Organism selector: *Homo sapiens* |
| **SuperPred** | https://prediction.charite.de/ | SMILES or name input |

See `reference/database_headers.md` for exact column differences between servers and versions.

## File naming convention: `Targets<PubChemCID>.csv`

> **Why PubChem CID, not CAS?**
> A single compound can have multiple CAS numbers registered across different chemical databases (racemate vs. enantiomer, anhydrous vs. hydrate, etc.). This creates silent mismatches when joining tables. The PubChem CID is a single, stable, unambiguous integer identifier.
>
> Example: Apigenin → `Targets5280443.csv`
>
> To find a PubChem CID: https://pubchem.ncbi.nlm.nih.gov/ → search by name → copy the CID integer.

```{r example-flag}
USE_EXAMPLE_DATA <- TRUE
EXAMPLE_DIR      <- "example_data"
```

```{r libraries}
library(tidyverse)
library(fs)
```

---

# Step 1 — Locate prediction files

```{r locate}
if (USE_EXAMPLE_DATA) {
  target_dir <- file.path(EXAMPLE_DIR, "04_targets")
} else {
  target_dir <- "04_target_prediction/raw"
  # Place your Targets<CID>.csv files in this folder
}

target_files <- dir_ls(target_dir, glob = "*.csv")
cat("Found", length(target_files), "prediction files\n")
```

---

# Step 2 — Header check

> ⚠️ **Always inspect headers before consolidating.** Column names differ by server
> and by version. Run this block on your first file to see what you have.

```{r check-headers}
if (length(target_files) > 0) {
  sample_df <- read_csv(target_files[1], show_col_types = FALSE, n_max = 3)
  cat("Headers in", basename(target_files[1]), ":\n")
  print(colnames(sample_df))
  print(head(sample_df, 3))
}
```

---

# Step 3 — Consolidate files

Choose the block that matches your server output format.

## Option A — SwissTargetPrediction format

> Expected headers: `Target Name`, `Gene Symbol`, `UniProt ID`, `Probability`, `Model accuracy`
>
> Probabilities are already in decimal (0–1) format.

```{r consolidate-stp}
safe_read_stp <- function(file) {
  df <- read_csv(file, show_col_types = FALSE)
  required <- c("Target Name", "Gene Symbol", "UniProt ID", "Probability")
  if (nrow(df) > 0 && all(required %in% colnames(df))) return(df)
  message("Skipping (unexpected format): ", basename(file))
  return(NULL)
}

targets_raw <- target_files %>%
  set_names(basename(.)) %>%
  map_dfr(safe_read_stp, .id = "file_name")

df_targets <- targets_raw %>%
  mutate(
    PubChemCID = as.integer(str_remove_all(file_name, "^Targets|\\.csv$"))
  ) %>%
  rename(UniProtID = `UniProt ID`) %>%
  filter(Probability > 0) %>%
  select(PubChemCID, `Target Name`, `Gene Symbol`, UniProtID,
         Probability, `Model accuracy`, -file_name)

cat("Total interactions stacked:", nrow(df_targets), "\n")
cat("Unique compounds:", n_distinct(df_targets$PubChemCID), "\n")
cat("Unique targets:", n_distinct(df_targets$UniProtID), "\n")
```

## Option B — SuperPred format

> Expected headers: `Target_name`, `uniprot_id`, `probability`, `model_acc`
>
> Probabilities may be 0–100 (percentage) — check and convert if needed.

```{r consolidate-superpred, eval=FALSE}
safe_read_superpred <- function(file) {
  df <- read_csv(file, show_col_types = FALSE)
  required <- c("Target_name", "uniprot_id", "probability")
  if (nrow(df) > 0 && all(required %in% colnames(df))) return(df)
  message("Skipping: ", basename(file))
  return(NULL)
}

targets_raw_sp <- target_files %>%
  set_names(basename(.)) %>%
  map_dfr(safe_read_superpred, .id = "file_name")

df_targets <- targets_raw_sp %>%
  mutate(
    PubChemCID  = as.integer(str_remove_all(file_name, "^Targets|\\.csv$")),
    Probability = if_else(probability > 1, probability / 100, probability)  # % → decimal
  ) %>%
  rename(
    `Target Name` = Target_name,
    `Gene Symbol` = gene_symbol,   # adjust if column name differs
    UniProtID     = uniprot_id,
    `Model accuracy` = model_acc
  ) %>%
  filter(Probability > 0) %>%
  select(PubChemCID, `Target Name`, `Gene Symbol`, UniProtID,
         Probability, `Model accuracy`)
```

---

# Step 4 — Attach compound metadata

```{r attach-metadata}
compounds_safe <- read_csv("03_toxicity_filtering/compounds_safe.csv",
                            show_col_types = FALSE)

df_targets_annotated <- df_targets %>%
  left_join(
    compounds_safe %>% select(Name, PubChemCID, SMILES),
    by = "PubChemCID"
  )

cat("Annotated interactions:", nrow(df_targets_annotated), "\n")
```

---

# Step 5 — Export

```{r export}
write_csv(df_targets_annotated, "04_target_prediction/df_targets.csv")
cat("Output: 04_target_prediction/df_targets.csv\n\n")

# Quick preview
df_targets_annotated %>%
  group_by(Name) %>%
  summarise(n_targets = n(), max_prob = max(Probability)) %>%
  arrange(desc(n_targets)) %>%
  print()
```

---

# Output

| File | Description |
|------|-------------|
| `04_target_prediction/raw/Targets<CID>.csv` | Individual prediction files (one per compound) |
| `04_target_prediction/df_targets.csv` | Consolidated compound–target interaction table |

→ Feeds into **Step 05 (Network Analysis)**
