---
title: "03 — Toxicity Filtering: ADMETlab Safety Screen"
author: "Edgard Uriel López Chablé"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Overview

Applies an early-stage toxicity screen using **ADMETlab 2.0 / 3.0** before committing to target prediction and docking. Discards compounds with unacceptable predicted safety profiles.

| Property | Meaning | Safe threshold |
|----------|---------|----------------|
| **Ames** | Mutagenicity | probability < 0.5 |
| **DILI** | Drug-induced liver injury | probability < 0.5 |
| **hERG** | Cardiac channel blockade | probability < 0.5 |

```{r example-flag}
USE_EXAMPLE_DATA <- TRUE
EXAMPLE_DIR      <- "example_data"
```

```{r libraries}
library(tidyverse)
```

---

# Step 1 — Load compounds from Step 02

```{r load}
compounds_filtered <- read_csv("02_adme_filtering/compounds_filtered.csv",
                                show_col_types = FALSE)
cat("Input compounds:", nrow(compounds_filtered), "\n")
```

---

# Step 2 — Generate ADMETlab input

```{r input}
for_admetlab <- compounds_filtered %>%
  select(SMILES, Name, PubChemCID)

write_csv(for_admetlab, "03_toxicity_filtering/for_admetlab.csv")
cat("ADMETlab input written:", nrow(for_admetlab), "compounds\n")
```

---

# Manual Step — Run ADMETlab

> **Action required (skip when using example data):**
>
> 1. Go to https://admetlab3.scbdd.com/ (v3) or https://admetmesh.scbdd.com/ (v2)
> 2. Paste the SMILES column or upload `for_admetlab.csv`
> 3. Run the prediction
> 4. Download the results CSV
> 5. Save as `03_toxicity_filtering/admetlab_results.csv`

---

# Step 3 — Header check

> ⚠️ Column names differ between ADMETlab versions. Inspect before filtering.
>
> | Property | v2 column | v3 column |
> |----------|-----------|-----------|
> | Input SMILES | `smiles` | `SMILES` |
> | Mutagenicity | `Ames` | `Ames_mutagenicity` |
> | Liver injury | `DILI` | `DILI` |
> | Cardiotoxicity | `hERG` | `hERG_blockers` |

```{r check-headers}
if (USE_EXAMPLE_DATA) {
  tox_raw <- read_csv(file.path(EXAMPLE_DIR, "03_admetlab_results.csv"),
                       show_col_types = FALSE)
} else {
  tox_raw <- read_csv("03_toxicity_filtering/admetlab_results.csv",
                       show_col_types = FALSE)
}

cat("ADMETlab columns detected:\n")
print(colnames(tox_raw))
```

---

# Step 4 — Rename columns to standard names

Adjust the `rename()` calls to match your version's actual headers.

```{r rename}
# Standard internal names used downstream
tox_results <- tox_raw %>%
  rename(
    smiles_tox = smiles,   # <- adjust if your column is named differently
    ames_prob  = Ames,     # <- adjust for v3: Ames_mutagenicity
    dili_prob  = DILI,
    herg_prob  = hERG      # <- adjust for v3: hERG_blockers
  )

cat("Column names after renaming:\n")
print(colnames(tox_results))
```

---

# Step 5 — Apply toxicity filter

```{r filter}
safe_compounds <- tox_results %>%
  filter(ames_prob < 0.5) %>%
  filter(dili_prob < 0.5) %>%
  filter(herg_prob < 0.5)

cat("Compounds passing toxicity filter:", nrow(safe_compounds), "\n")
cat("Compounds removed (predicted toxic):",
    nrow(tox_results) - nrow(safe_compounds), "\n")
```

---

# Step 6 — Merge with compound metadata and export

```{r merge-export}
compounds_safe <- compounds_filtered %>%
  inner_join(safe_compounds, by = c("SMILES" = "smiles_tox"))

write_csv(compounds_safe, "03_toxicity_filtering/compounds_safe.csv")
cat("Output: 03_toxicity_filtering/compounds_safe.csv —",
    nrow(compounds_safe), "compounds\n")
```

---

# Filtering Summary

```{r summary}
cat("=== Filtering Summary ===\n")
cat("Step 01 (cleaning):        ", n_distinct(compounds_safe$Name), "compounds in\n")
cat("Step 02 (ADME filter):     ", nrow(compounds_filtered), "compounds\n")
cat("Step 03 (toxicity filter): ", nrow(compounds_safe), "compounds\n")
```

---

# Output

| File | Description |
|------|-------------|
| `03_toxicity_filtering/for_admetlab.csv` | SMILES list for ADMETlab submission |
| `03_toxicity_filtering/compounds_safe.csv` | Compounds passing ADME + toxicity screens |

→ Feeds into **Step 04 (Target Prediction)**
