---
title: "00 — Setup: Package Installation & Project Scaffold"
author: "Edgard Uriel López Chablé"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Overview

This script handles two things:

1. **Package installation** — all R/Bioconductor packages required for the full pipeline.
2. **Project folder scaffold** — creates the recommended directory structure for a new analysis project.

Run this file **once** before starting a new project. You do not need to re-run it in subsequent sessions.

---

# 1. Package Installation

## CRAN packages

```{r cran}
install.packages(c(
  "tidyverse",      # Core data manipulation and plotting
  "webchem",        # Chemical identifier lookup (PubChem, CIR)
  "igraph",         # Graph construction and analysis
  "ggraph",         # Grammar-of-graphics network visualization
  "ggrepel",        # Non-overlapping text labels
  "ggVennDiagram",  # Venn diagram visualization
  "pheatmap",       # Heatmap with hierarchical clustering
  "UpSetR",         # UpSet intersection plots
  "RColorBrewer",   # Color palettes
  "magick",         # Image processing (for pathview PDF compilation)
  "fs",             # File system operations
  "readxl",         # Read Excel files
  "openxlsx"        # Write Excel files
))
```

## Bioconductor packages

```{r bioc}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "clusterProfiler",  # GO, KEGG enrichment analysis
  "org.Hs.eg.db",     # Human gene annotation database
  "ReactomePA",       # Reactome pathway enrichment
  "pathview",         # KEGG pathway map visualization
  "enrichplot",       # Enrichment visualization (dotplot, cnetplot, etc.)
  "GOplot",           # GOChord and other GO visualization
  "STRINGdb",         # Protein-protein interaction data from STRING
  "reactome.db"       # Reactome database for annotation
))
```

## Verify installation

```{r verify}
required_pkgs <- c(
  "tidyverse", "webchem", "igraph", "ggraph", "ggrepel", "ggVennDiagram",
  "pheatmap", "UpSetR", "RColorBrewer", "magick", "fs",
  "clusterProfiler", "org.Hs.eg.db", "ReactomePA", "pathview",
  "enrichplot", "GOplot", "STRINGdb"
)

missing <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]

if (length(missing) == 0) {
  cat("All packages installed successfully!\n")
} else {
  cat("Missing packages:\n")
  print(missing)
}
```

---

# 2. Project Folder Scaffold

This creates the recommended directory structure for a new analysis. Adjust `project_root` to your desired path.

```{r scaffold}
library(fs)

# ── Set your project root ──────────────────────────────────────────────────────
project_root <- "MyPlant_NetworkPharmacology"   # <- Change this

# ── Folder structure ───────────────────────────────────────────────────────────
folders <- c(
  # Step 01: Data cleaning inputs and outputs
  "01_data_cleaning",

  # Step 02: ADME
  "02_adme_filtering/input",    # <- place Para_SwissADME.txt here
  "02_adme_filtering/output",   # <- place swissadme.csv here after download

  # Step 03: Toxicity
  "03_toxicity_filtering/input",   # <- place SMILES list for ADMETlab here
  "03_toxicity_filtering/output",  # <- place admetlab_results.csv here

  # Step 04: Target prediction
  # Name files as: Targets<PubChemCID>.csv  (e.g., Targets2723949.csv)
  # One file per compound, downloaded from SuperPred or SwissTargetPrediction
  "04_target_prediction/raw",      # <- place individual Targets*.csv here
  "04_target_prediction/output",   # <- consolidated df_targets.csv goes here

  # Step 05: Network analysis
  "05_network_analysis/genecards", # <- place GeneCards*.csv files here
  "05_network_analysis/output",

  # Step 06: Pre-docking analysis
  "06_predocking_analysis/output"
)

dir_create(path(project_root, folders), recurse = TRUE)

cat("Project scaffold created at:", project_root, "\n")
dir_tree(project_root)
```

---

# Notes

- Keep **raw data files unchanged** in their input folders. Never overwrite them.
- All generated outputs are written to `output/` subfolders.
- The `reference/` folder in this cookbook contains notes on expected file headers from each database.
- The PubChem CID for any compound can be found at: https://pubchem.ncbi.nlm.nih.gov/
