---
title: "02 — ADME Filtering: Lipinski Rule-of-Five via SwissADME"
author: "Edgard Uriel López Chablé"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Overview

Filters compounds by oral drug-likeness using **SwissADME** (https://www.swissadme.ch/).

The **Lipinski Rule-of-Five** predicts oral bioavailability:

| Property | Threshold |
|----------|-----------|
| Molecular weight | ≤ 500 Da |
| LogP (lipophilicity) | ≤ 5 |
| H-bond donors | ≤ 5 |
| H-bond acceptors | ≤ 10 |

We allow ≤ 1 violation (permissive, appropriate for natural product extracts). Additionally, we require GI absorption to be present.

```{r example-flag}
USE_EXAMPLE_DATA <- TRUE
EXAMPLE_DIR      <- "example_data"
```

```{r libraries}
library(tidyverse)
```

---

# Step 1 — Load compounds from Step 01

```{r load}
compounds_ready <- read_csv("01_data_cleaning/compounds_ready.csv",
                             show_col_types = FALSE)
cat("Input compounds:", nrow(compounds_ready), "\n")
```

---

# Step 2 — Generate SwissADME input file

SwissADME accepts a plain-text file: one entry per line, `SMILES IDENTIFIER` (space-separated, no header). We use the **PubChem CID** as identifier — it is a unique, stable integer that makes the merge back unambiguous.

```{r swissadme-input}
for_swissadme <- compounds_ready %>%
  select(SMILES, PubChemCID)

write.table(for_swissadme,
            file      = "02_adme_filtering/for_swissadme.txt",
            sep       = " ",
            row.names = FALSE,
            col.names = FALSE,
            quote     = FALSE)

cat("SwissADME input written:", nrow(for_swissadme), "compounds\n")
cat("File: 02_adme_filtering/for_swissadme.txt\n")
```

---

# Manual Step — Run SwissADME

> **Action required (skip when using example data):**
>
> 1. Go to https://www.swissadme.ch/
> 2. Upload or paste the contents of `for_swissadme.txt`
> 3. Run the analysis
> 4. Download the CSV results
> 5. Save as `02_adme_filtering/swissadme_results.csv`

---

# Step 3 — Load SwissADME results

## ⚠️ Header check — always run this first

SwissADME occasionally renames columns between versions. The essential columns for this pipeline are:

| Column name | Meaning |
|-------------|---------|
| `Molecule` | Identifier submitted (= PubChemCID) |
| `Lipinski..violations` | Integer 0–4 |
| `GI.absorption` | `"High"` or `"Low"` |

Run `colnames(read.csv("swissadme_results.csv"))` to verify before proceeding.

```{r load-swissadme}
if (USE_EXAMPLE_DATA) {
  swissadme_raw <- read_csv(file.path(EXAMPLE_DIR, "02_swissadme.csv"),
                             show_col_types = FALSE)
} else {
  swissadme_raw <- read_csv("02_adme_filtering/swissadme_results.csv",
                             show_col_types = FALSE)
}

# Quick header check
cat("SwissADME columns detected:\n")
print(colnames(swissadme_raw))
```

```{r merge}
adme_results <- swissadme_raw %>%
  rename(PubChemCID = Molecule) %>%
  mutate(PubChemCID = as.integer(PubChemCID))

compounds_with_adme <- compounds_ready %>%
  left_join(adme_results, by = "PubChemCID")

cat("Compounds after merge:", nrow(compounds_with_adme), "\n")
```

---

# Step 4 — Apply Lipinski filter

```{r filter}
# Adjust thresholds to your project:
#   Natural products (permissive): Lipinski..violations <= 1, GI.absorption %in% c("Yes","High")
#   Strict drug-likeness:          Lipinski..violations == 0, GI.absorption == "High"

compounds_filtered <- compounds_with_adme %>%
  filter(Lipinski..violations <= 1) %>%
  filter(GI.absorption %in% c("High", "Yes"))

cat("Compounds passing ADME filter:", nrow(compounds_filtered), "\n")
cat("Compounds removed:            ", nrow(compounds_with_adme) - nrow(compounds_filtered), "\n")
```

---

# Step 5 — Export

```{r export}
write_csv(compounds_filtered, "02_adme_filtering/compounds_filtered.csv")
cat("Output: 02_adme_filtering/compounds_filtered.csv\n")
print(compounds_filtered %>% select(Name, PubChemCID, Lipinski..violations, GI.absorption))
```

---

# Output

| File | Description |
|------|-------------|
| `02_adme_filtering/for_swissadme.txt` | Input file for SwissADME |
| `02_adme_filtering/compounds_filtered.csv` | Compounds passing Lipinski filter |

→ Feeds into **Step 03 (Toxicity Filtering)**
